The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?

### 2023-02-03 outputs/paper/paper.qmd
Paper updates
* Included a summary of findings in the introduction
* Fixed formatting issues

A city's emergency medical services (EMS) are integral to its health care system. The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance. These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires. Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour]. This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].

@@

Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival. This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments. Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service. \newpage

@@

The Toronto Paramedic Services should include more comprehensive metrics in the data sets they provide to Open Data Toronto. As revealed in this paper, the service's current data sets lack vital information on case-by-case response times and patient outcomes. Where permitted, the service should also include the type of medical emergency addressed by paramedics. The inclusion of these expanded data points would allow for a more meaningful analysis of their data. \newpage

>>>>

A city's emergency medical services (EMS) are integral to its health care system. The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance. These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires. Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour]. This "golden hour" is the foundation of all emergency medical responses and requires medical personnel to arrive on scene in under 10 minutes [@cite_golden_hour_10_mins].

@@

Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival. This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments. Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service.

The findings indicated a growing trend in the number of emergency medical calls, an increase of approximately 10,000 per year since 2014 [@cite_emergency_care_data]. Despite this increase, the Toronto Paramedic Services met and often exceeded the response times set by both the Province of Ontario and the City of Toronto. The paper also revealed a significant lack of paramedic metrics from the data sets available on Open Data Toronto. Without metrics on case-by-case response times and patient outcomes, other sources were required to inform data analysis. Overall, the response time of Toronto's paramedics contributed to stroke and mortality rates lower than Ontario's average, giving the city's inhabitants greater chances of survival when suffering from life-threatening injuries. \newpage

@@

The Toronto Paramedic Services should include more comprehensive metrics in the data sets they provide to Open Data Toronto. As revealed in this paper, the service's current data sets lack vital information on case-by-case response times and patient outcomes. Where permitted, the service should also include the type of medical emergency addressed by paramedics. The inclusion of these expanded data points would allow for a more meaningful analysis. \newpage
